A Systematic Literature Review on Extensions to Role-Based Access Control

### 2013-05-15 paper/sections/abstract.tex
Removing duplicate paper and updating total counts.

\noindent
{\bf Context}: Since the US National Institute of Standards and Technology (NIST) proposed Role-Based Access Control (RBAC) as documented in the RBAC Reference Model in the late 1990's, domain-specific extensions have been proposed.  For example, the mobile domain has surfaced a need for permission granting and role activation.
\\
\\
\noindent
{\bf Aim}: The goal of our work is to aid practitioners and researchers in choosing an RBAC extension model, and in understanding
how RBAC extension models are evaluated by providing an assessment of the state of the RBAC extension models.
\\
\\
\noindent
{\bf Method}: We performed a systematic literature review that began with 1,716 paper of which 28 were deemed primary sources for inclusion as the RBAC extension models. We established a set of extension categories, examined the state of the art in evaluations of the RBAC extension models, and categorized the motivations that have led to the RBAC extension models.
\\
\\
\noindent
{\bf Results}: We classified the RBAC extension models into eight extension categories: Constraint, Context, 
Organization, Privacy, Task, Spatio-Temporal, Spatial, and Temporal. only 8 of the 28 papers provided an implementation of their model in the form of an enterprise application or prototype. The primary domains 
that inspired extensions were the medical domain with 9 of the 28 models, enterprise workflows with 5 of 28, and mobile computing with 5 of 28.
\\
\\
\noindent
{\bf Conclusions}: All of the eight RBAC extension categories deal with context. 
Our literature review shows that the state of evaluation of RBAC model evaluation is limited with 1
paper doing an evaluation comparing itself to the RBAC Reference Model, 8 presenting example 
scenarios of their model in action, and 12 of the 28 models lacking an evaluation that we could discern. The magnitude of extensions to the RBAC standard suggests a revised standard may be beneficial.

>>>>

\noindent
{\bf Context}: Since the US National Institute of Standards and Technology (NIST) proposed Role-Based Access Control (RBAC) as documented in the RBAC Reference Model in the late 1990's, domain-specific extensions have been proposed.  For example, the mobile domain has surfaced a need for permission granting and role activation.
\\
\\
\noindent
{\bf Aim}: The goal of our work is to aid practitioners and researchers in choosing an RBAC extension model, and in understanding
how RBAC extension models are evaluated by providing an assessment of the state of the RBAC extension models.
\\
\\
\noindent
{\bf Method}: We performed a systematic literature review that began with 1,716 paper of which 27 were deemed primary sources for inclusion as the RBAC extension models. We established a set of extension categories, examined the state of the art in evaluations of the RBAC extension models, and categorized the motivations that have led to the RBAC extension models.
\\
\\
\noindent
{\bf Results}: We classified the RBAC extension models into eight extension categories: Constraint, Context, 
Organization, Privacy, Task, Spatio-Temporal, Spatial, and Temporal. only 8 of the 27 papers provided an implementation of their model in the form of an enterprise application or prototype. The primary domains 
that inspired extensions were the medical domain with 9 of the 27 models, enterprise workflows with 5 of 27, and mobile computing with 5 of 27.
\\
\\
\noindent
{\bf Conclusions}: All of the eight RBAC extension categories deal with context. 
Our literature review shows that the state of evaluation of RBAC model evaluation is limited with 1
paper doing an evaluation comparing itself to the RBAC Reference Model, 8 presenting example 
scenarios of their model in action, and 12 of the 27 models lacking an evaluation that we could discern. The magnitude of extensions to the RBAC standard suggests a revised standard may be beneficial.

### 2013-05-15 paper/sections/analysis.tex
Updates to address dr. williams' comments
Removing duplicate paper and updating total counts.

We first describe motivations described in the 28 papers surveyed.

@@

We classify motivations behind RBAC extension models to these three categories:

@@

%\begin{table}
%\centering
%\caption{Motivation categories found and the count of primary sources}
%\begin{tabular}{ | c | c | }
%\cline{1-2}
%
%\textbf{Motivation Category} & \textbf{Paper Count} \\ \cline{1-2}
%Addition & 23 \\ \cline{1-2}
%Modification & 1 \\ \cline{1-2}
%Combination & 4 \\
%
%\cline{1-2}
%\end{tabular}
%\label{tab:motivations}
%\end{table}
%
%Table \ref{tab:motivations} shows the breakdown of motivations found within the primary sources.
%The result shows that 23 papers out of 28 papers are classified into ``Addition'' category. Only
%one paper and four papers are classified into ``Modification'' category and ``Combination'' category, respectively.

@@

Table \ref{tab:implementations} shows the breakdown of implementations found within the primary sources.
Of the 28 papers surveyed, 4 papers~\cite{zhou2007team,bao08:role, cholewka00:acontext-sensitive,huang06:pervasive} were simply prototypes developed by the authors whereas 4 papers~\cite{jian2008extended, aich09:role, yao2008task, motta03:contextual} were claimed to be implemented within a real
system. The remaining 20 papers provide no
mention of an implementation or prototype.

@@

The 28 primary sources were examined for evidence that evaluations of the proposed model were presented by the model authors.
Further, we identified a set of evaluation types found within  each of the primary sources and provide below a list of the 
evaluation types and which primary sources provided which type.  In some cases a single primary source provided multiple evaluation
types.

@@

The predominant domain for which extension models have been generated for is that of the medical domain with 9 of 28 mentioning scenarios or requirements of that industry.
Mobile computing and enterprise workflows were each represented by five papers claiming to be influenced by the requirements for access control within these domains. The final set
of domains was pervasive computing environments and large-scale organizations with three each and web applications with one.  There were three papers~\cite{jian2008extended, aich07:STARBAC, haibo2005context} without any direct mention of a domain since Aich et al. \cite{aich09:role} fall under both the medical domain and mobile computing.

@@

Since NIST proposed RBAC standard using propositional logic, researchers describe extended RBAC models using propositional logic. Given context, propositional logic is used to evaluate true or false based on specified rules and access control logic. As access control is typically evaluated to either true or false based on predicates, propositional logic is sufficient to describe key ideas and definitions. We found that 28 papers of RBAC extensions use propositional logic to describe its extended model.
However, propositional logic has limitations. While this logic is simple, this logic does not support for reasoning about RBAC, which helps
reduce the administrative complexity of associations such as user- role associations.
In order to support for reasoning of access control, one may describe the RBAC extended models in first-order logic. 
Given an RBAC extension model, Samuel et al.~\cite{samuel07:spatio-temporal} proposed verification of the model using a specification language, which is based first-order logic. This logic is sufficient to model RBAC and extended RBAC for reasoning. Moreover, this logic supports for concise and elegant formulation of the Reference Model and its relation.  First-Order logic is expressive enough to concisely represent access control systems. First-Order logic uses relations, variables, and quantifiers.

>>>>

We first describe motivations described in the 27 papers surveyed.

@@

We classify motivations behind the RBAC extension models to these three categories:

@@

%\begin{table}
%\centering
%\caption{Motivation categories found and the count of primary sources}
%\begin{tabular}{ | c | c | }
%\cline{1-2}
%
%\textbf{Motivation Category} & \textbf{Paper Count} \\ \cline{1-2}
%Addition & 23 \\ \cline{1-2}
%Modification & 1 \\ \cline{1-2}
%Combination & 4 \\
%
%\cline{1-2}
%\end{tabular}
%\label{tab:motivations}
%\end{table}
%
%Table \ref{tab:motivations} shows the breakdown of motivations found within the primary sources.
%The result shows that 23 papers out of 27 papers are classified into ``Addition'' category. Only
%one paper and four papers are classified into ``Modification'' category and ``Combination'' category, respectively.

@@

Table \ref{tab:implementations} shows the breakdown of implementations found within the primary sources.
Of the 27 papers surveyed, 4 papers~\cite{zhou2007team,bao08:role, cholewka00:acontext-sensitive,huang06:pervasive} were simply prototypes developed by the authors whereas 4 papers~\cite{jian2008extended, aich09:role, yao2008task, motta03:contextual} were claimed to be implemented within a real
system. The remaining 20 papers provide no
mention of an implementation or prototype.

@@

The 27 primary sources were examined for evidence that evaluations of the proposed model were presented by the model authors.
Further, we identified a set of evaluation types found within  each of the primary sources and provide below a list of the 
evaluation types and which primary sources provided which type.  In some cases a single primary source provided multiple evaluation
types.

@@

The predominant domain for which extension models have been generated for is that of the medical domain with 9 of 27 mentioning scenarios or requirements of that industry.
Mobile computing and enterprise workflows were each represented by five papers claiming to be influenced by the requirements for access control within these domains. The final set
of domains was pervasive computing environments and large-scale organizations with three each and web applications with one.  There were three papers~\cite{jian2008extended, aich07:STARBAC, haibo2005context} without any direct mention of a domain since Aich et al. \cite{aich09:role} fall under both the medical domain and mobile computing.

@@

Since NIST proposed RBAC standard using propositional logic, researchers describe extended RBAC models using propositional logic. Given context, propositional logic is used to evaluate true or false based on specified rules and access control logic. As access control is typically evaluated to either true or false based on predicates, propositional logic is sufficient to describe key ideas and definitions. We found that 27 papers of RBAC extensions use propositional logic to describe its extended model.
However, propositional logic has limitations. While this logic is simple, this logic does not support for reasoning about RBAC, which helps
reduce the administrative complexity of associations such as user- role associations.
In order to support for reasoning of access control, one may describe the RBAC extended models in first-order logic. 
Given an RBAC extension model, Samuel et al.~\cite{samuel07:spatio-temporal} proposed verification of the model using a specification language, which is based first-order logic. This logic is sufficient to model RBAC and extended RBAC for reasoning. Moreover, this logic supports for concise and elegant formulation of the Reference Model and its relation.  First-Order logic is expressive enough to concisely represent access control systems. First-Order logic uses relations, variables, and quantifiers.
